Normalization by Evaluation for Sized Dependent Types

### 2017-02-28 nbe/icfp17.tex
[ ICFP 17 ] Another read of section 4.

The \emph{read back} phase of NbE \cite{gregoireLeroy:icfp02}
transforms a reified value $d$ into a normal form $v$.  It is
specified via an inductively defined relation $\R k d v$ and several auxiliary relations.
The number $k$, will be instantiated by the length of the context
$\Gamma$ later.  It allows us to transform a de Bruijn level $l$ into a de
Bruijn index $i$, via the law $i + l + 1 = k$.  At this point, we do
not ensure that the $k$ is large enough to accommodate the de Bruijn
levels in $d$.  Levels $l \geq k$ which are to big will simply mapped
to de Bruijn index $0$.  The correct $k$ is later ensured by our
logical relation (Section~\ref{sec:soundnbe}).
% Judgements.
% \[
% \begin{array}{l@{\qquad}l}
%    \R k d  v     & \mbox{} \\
%    \RTy k T  V  & \mbox{} \\
%    \RNat k t  v & \mbox{} \\
%    \RNe k t  m  & \mbox{} \\
% \end{array}
% \]
Even though read back operates on values in practice, formally it is
simply defined on expressions.

@@

A type $T$ will be interpreted as a partial equivalence relation (PER) $\A$ on terms, \ie, a relation which is symmetric and transitive. The domain $\dom(\A)$ of the relation can be thought of as the set of terms which define the extension of the type; on $\dom(\A) = \{a \mid \exists a'.\ (a,a') \in \A \}$ the relation $\A$ is in fact an equivalence relation.  We write $a = a' \in \A$ for relatedness in $\A$ and $a \in \A$ if $a \in \dom(\A)$.

@@

%Neutrals in the value world $\D$.
PER $\NEE$ interprets all neutral types.
\[
\fbox{$t = t' \in \NEE$} \defas
  t \evalsto \up T n \mand t' \evalsto \up{T'}{n'} \mand
  n = n' \in \NE
% \fbox{$\NEE$} \defas \{ (\up T n, \up{T'}{n'})
%   % \msuchthat T \evalsto N \mand T' \evalsto N' \mand
%   \mid n = n' \in \NE \}
\]

@@

Reflection and reification perform $\eta$-expansion such that we arrive at an $\eta$-long $\beta$-normal form.  To perform the $\eta$-expansion, not the precise type is needed, just the approximate shape, in particular, whether it is a function type (do expand) or a base type (do not expand).
For the logical framework, the shape of a dependent type is just its
underlying simple type \cite{harperPfenning:equivalenceLF}. However,
in the presence of universes and large eliminations, there is no
underlying simple type.  Of course, we can take a type as its own
shape, but we want at least that $\Nat \alpha$ and $\Nat \beta$ have
the same shape even for different $\alpha, \beta$.
Also all neutral types can be summarized under a single shape.
We make our intuition precise by defining a relation $T \shape S$
between type values, to express that $S$ is a possible shape of type $T$.

@@

Further, it is not true that equal types make equally good shapes.  We do not have that
$T \shape S$ and $S = S' \in \SET[\ell]$ imply $T \shape S'$.  This property fails for function types.
Given $\piT U T \shape \piT R S$ and $\piT R S = \piT {R'}{S'} \in \SET[\ell]$ we would need to show that $T[u] \shape S'[u']$ for all $u = u' \in U$, but we only have $S[u] = S'[u'] \in \SET[\ell]$ for all $u = u' \in R$, thus the induction does not go through.  The fact that $U \shape R$ does not give us a handle on their inhabitants, we would need $U \leq R \in \TYPE$.
It is possible to construct an actual counterexample, using
$\piT {R'} S' =  \funT x {\Nat 0} {F\,x}$ from above and $\piT R S = \funT x {\Nat 0} {G\,x}$
such that $G$ is defined on all of $\Nat \infty$ but agrees with $F$ only on $x \in \Nat 0$.
Then $\piT U T = \funT x {\Nat \infty} {G\,x}$ gives the desired counterexample.

@@

\nextcase $\alpha = \infty$.
Note that $a = a' \in \Nat \alpha$ for some $\alpha < \infty$ by Lemma~\ref{lem:limit}.
By induction hypothesis, $g = g' \in B\,\alpha\,a$.  Since
$B\,\alpha\,a \leq B\,\infty\,a$ by assumption, the goal
$g = g' \in B\,\infty\,a$ follows by subsumption.
\qed
\end{caselist}
\end{proof*}

@@

From the fundamental theorem, we harvest completeness of NbE in this
section, \ie, we show that definitionally equal terms have the same
normal form.
We may write simply $\Gamma$ for the length $|\Gamma|$ where there is no
danger of confusion, for instance in de Bruijn level $\var\Gamma$ or
in read back $\tR_{\Gamma}$.
%We write $\var\Gamma$ for $\var{|\Gamma|}$.
We define the \emph{identity environment} \fbox{$\rho_\Gamma$} by induction on $\Gamma$,
setting $\rho_{\cempty} = \sempty$ and
$\rho_{\eext \Gamma \Size} = (\rho_\Gamma, \var\Gamma)$ and
$\rho_{\cext \Gamma T} = (\rho_\Gamma, \up{T\rho_\Gamma} \var\Gamma)$.
% \[
% \begin{array}{lll}
%   \rho_{\cext \Gamma T} & = & (\rho_\Gamma, \up{T\rho_\Gamma} \var\Gamma) \\
%   \rho_{\eext \Gamma \Size} & = & (\rho_\Gamma, \var\Gamma) \\
% \end{array}
% \]
\begin{lemma}[Identity environment]
  If\/ $\der \Gamma$ then
  $\rho_\Gamma \in \Gamma$.
  % $\rho_\Gamma = \rho_\Gamma \eeq \rho_\Gamma \in \Gamma$
\end{lemma}
\begin{proof*}
By induction on $\der \Gamma$.
\begin{caselist}

>>>>

The \emph{read back} phase of NbE \cite{gregoireLeroy:icfp02}
transforms a reified value $d$ into a normal form $v$.  It is
specified via an inductively defined relation $\R k d v$ and several auxiliary relations.
The number $k$, will be instantiated by the length of the context
$\Gamma$ later.  It allows us to transform a de Bruijn level $l$ into a de
Bruijn index $i$, via the law $i + l + 1 = k$.  At this point, we do
not ensure that the $k$ is large enough to accommodate the de Bruijn
levels in $d$.  Levels $l \geq k$ which are to big will simply be mapped
to de Bruijn index $0$.  The correct $k$ is later ensured by our
logical relation (Section~\ref{sec:soundnbe}).
% Judgements.
% \[
% \begin{array}{l@{\qquad}l}
%    \R k d  v     & \mbox{} \\
%    \RTy k T  V  & \mbox{} \\
%    \RNat k t  v & \mbox{} \\
%    \RNe k t  m  & \mbox{} \\
% \end{array}
% \]
Even though read back operates on values in practice, formally it is
defined on expressions.

@@

A type $T$ will be interpreted as a partial equivalence relation (PER) $\A$ on terms, \ie, a relation which is symmetric and transitive. The domain $\dom(\A)$ of the relation can be thought of as the set of terms which denotes the extension of the type; on $\dom(\A) = \{a \mid \exists a'.\ (a,a') \in \A \}$ the relation $\A$ is in fact an equivalence relation.  We write $a = a' \in \A$ for relatedness in $\A$ and $a \in \A$ if $a \in \dom(\A)$.

@@

%Neutrals in the value world $\D$.
PER $\NEE$ interprets all neutral types.
\[
\fbox{$t = t' \in \NEE$} \defas
  t \evalsto \up T n \mand t' \evalsto \up{T'}{n'} \mand
  n = n' \in \NE
  .
% \fbox{$\NEE$} \defas \{ (\up T n, \up{T'}{n'})
%   % \msuchthat T \evalsto N \mand T' \evalsto N' \mand
%   \mid n = n' \in \NE \}
\]

@@

Reflection and reification perform $\eta$-expansion such that we arrive at an $\eta$-long $\beta$-normal form.  To perform the $\eta$-expansion, not the precise type is needed, just the approximate shape, in particular, whether it is a function type (do expand) or a base type (do not expand).
For the logical framework, the shape of a dependent type is just its
underlying simple type \cite{harperPfenning:equivalenceLF}. However,
in the presence of universes and large eliminations, there is no
underlying simple type.  Of course, we can take a type as its own
shape, but we want at least that $\Nat \alpha$ and $\Nat \beta$ have
the same shape even for different $\alpha, \beta$.
Also all neutral types can be summarized under a single shape.

We make our intuition precise by defining a relation $T \shape S$
between type values, to express that $S$ is a possible shape of type
$T$.  The asymmetry of this relation stems from the case for function
types.  At function types
$\piT A B \shape \piT R S$, we take $S$ to be a family over domain
$A$, not $R$!  We cannot take $R$ since we have to compare families $B$ and $S$
at a common domain, and $A$ and $R$ are not equal.

@@

Further, it is not true that equal types make equally good shapes.  We do not have that
$T \shape S$ and $S = S' \in \SET[\ell]$ imply $T \shape S'$.  This property fails for function types.
Given $\piT U T \shape \piT R S$ and $\piT R S = \piT {R'}{S'} \in \SET[\ell]$ we would need to show that $T[u] \shape S'[u']$ for all $u = u' \in U$, but we only have $S[u] = S'[u'] \in \SET[\ell]$ for all $u = u' \in R$, thus the induction does not go through.  The fact that $U \shape R$ does not give us a handle on their inhabitants, we would need a stronger relation such as $U \leq R \in \TYPE$.
It is possible to construct an actual counterexample, using
$\piT {R'} S' =  \funT x {\Nat 0} {F\,x}$ from above and $\piT R S = \funT x {\Nat 0} {G\,x}$
such that $G$ is defined on all of $\Nat \infty$ but agrees with $F$ only on $x \in \Nat 0$.
Then $\piT U T = \funT x {\Nat \infty} {G\,x}$ gives the desired counterexample.

@@

\nextcase $\alpha = \infty$.
Note that $a = a' \in \Nat \alpha$ for some $\alpha < \infty$ by Lemma~\ref{lem:limit}.
By induction hypothesis, $g = g' \in B\,\alpha\,a$.  Since
$B\,\alpha\,a \leq B\,\infty\,a$ by admissibility of $B$, the goal
$g = g' \in B\,\infty\,a$ follows by subsumption.
\qed
\end{caselist}
\end{proof*}

@@

From the fundamental theorem, we harvest completeness of NbE in this
section, \ie, we show that definitionally equal terms have the same
normal form.
We may write simply $\Gamma$ for its length $|\Gamma|$ when there is no
danger of confusion, for instance in de Bruijn level $\var\Gamma$ or
in read back $\tR_{\Gamma}$.
%We write $\var\Gamma$ for $\var{|\Gamma|}$.
We define the \emph{identity environment} \fbox{$\rho_\Gamma$} by induction on $\Gamma$,
setting $\rho_{\cempty} = \sempty$ and
$\rho_{\eext \Gamma \Size} = (\rho_\Gamma, \var\Gamma)$ and
$\rho_{\cext \Gamma T} = (\rho_\Gamma, \up{T\rho_\Gamma} \var\Gamma)$.
% \[
% \begin{array}{lll}
%   \rho_{\cext \Gamma T} & = & (\rho_\Gamma, \up{T\rho_\Gamma} \var\Gamma) \\
%   \rho_{\eext \Gamma \Size} & = & (\rho_\Gamma, \var\Gamma) \\
% \end{array}
% \]
\begin{lemma}[Identity environment]
  If\/ $\der \Gamma$ then
  $\rho_\Gamma \in \Gamma$.
  % $\rho_\Gamma = \rho_\Gamma \eeq \rho_\Gamma \in \Gamma$
\end{lemma}
\begin{proof*}
By induction on $\der \Gamma$.
\begin{caselist}
